Algebraic Filtering of Surfaces from 3D Medical Images with Julia

### 2020-04-26 src/design.tex

\subsubsection*{Boundary operator }\label{sec:boundary-operator}
For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.

>>>>

\subsubsection*{Boundary operator }\label{sec:boundary-operator}
For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~
\cite{TSAS}, 
% \cite{DiCarlo2014}, 
and inlined in the generated boundary extraction code.

### 2020-04-26 src/implementation.tex

\section{Julia implementation}\label{sec:implementation}

The computer code is implemented in Julia language~\cite{} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.

@@

\paragraph{Workflow setup} The functions in this preliminary step include:
\begin{enumerate}

\item input of 3D medical image $\mathcal{I}$ dimensions $\ell_1, \ell_2, \ell_3$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = (1,2,\ldots,\ell_k)$;

@@

\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $\mathcal{B}$ (bricks); defaults to: $\emph{size}=64$; hence the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$. 
Hence default value: $n=256$, for images $\mathcal{I} = 512\times 512\times 256$;

\item computation of a Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the average $[\partial_B]$ (for $\emph{size}$ = 64) is about 45 MB;

\item creation of either a local or distributed channel to implement a producer/consumer model of parallel/distributed computation, depending on available resources;

\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (workers) using the macro \texttt{@eveywhere}.

@@

The memory occupancy of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.  
With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a storage size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.

\begin{figure}[htbp]
\includegraphics[width=0.99\textwidth]{figs/schema_horizontal.pdf} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\caption{Workflow of LAR-SURF algorithm}
\label{fig:schema}
\end{figure}

With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB. The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.

\paragraph{Job enqueuing} Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication. Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete. The single work items of this stage follow:
\begin{enumerate}

@@

\paragraph{3-Chain encoding} 
The interesting part of the \emph{Image} $\mathcal{I}$ is called \emph{Segment} $\mathcal{S}$. The goal of the whole \emph{workflow} is to extract a \emph{boundary model} of $\mathcal{S}$ from $\mathcal{I}$. The portion of $\mathcal{S}$ inside $\mathcal{B}$, will be denoted as $\mathcal{S}(B)$.
Each block $\mathcal{B}$ of the 3D image must by converted into the \emph{coordinate representation} of a vector $\nu\in C_3$ in the linear space  of 3-chains.

@@

\paragraph{SpMM Multiplication} 
According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications. Since we have 256 such jobs (one multiplication per block) to perform in the standard setting of the algorithm\footnote{
Size of block $64^3$; size of image $512^2\times 256$.
}, or more in case of either smaller blocks or image greater than the standard one, this stage must be evidently parallelized and carefully tuned, possibly by using the GPU, if available.
\begin{enumerate}

\item Various multiplication algorithms are being experimented, using several packages for sparse linear algebra and/or custom implementations;

@@

\paragraph{2-Chain decoding} 
Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the \emph{coordinate representation} of the \emph{boundary vector} $\sigma\in C_2$.  The inverse of the coding algorithm is executed in the present stage.  This process can also be partially superimposed in time with the previous ones, depending on the size of the memory buffers used to feed the CPU cores or the GPUs and get their results. Some elementary steps follow:

@@

\paragraph{Assembling and artifact filtering} 
The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the image block $\mathcal{B}$.  In other words, \emph{all quads} of \textbf{each job} are now given in the \textbf{same} \emph{local coordinates}.  Besides to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $\mathcal{B}$ such that $B \cap \mathcal{I} \not= \emptyset$:
\begin{enumerate}

@@

\item remove both instances of \emph{double quads} generated by \texttt{Lar} software at the block boundaries (see Figure~\ref{fig:blockboundary}). They are artifacts generated by the decomposition of the whole image into a number of blocks of tractable size.

\item 
a smart strategy of removal of such artifacts may be used, which does not require any sorting nor searching on the assembled array of quads. It will consist in arranging each block with all three dimensions decreased-increased by one, so that each 2-adjacent pair of blocks will be covering each other for a full side extent of blocks of depth one. The details of this \emph{artifact filtering} are elucidated in Section~\ref{sec:covering}.

@@

\paragraph{Smoothing} 
The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries. Anyway, the Taubin smoothing~\cite{} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.  Some appropriate set workers may so be assigne the task of generating iteratively a new position for the vertices they take cure of. In particular, we have:
\begin{enumerate}

\item Job enqueuing, by writing sets of integers (global linear indices of vertices) in array buffers of appropriate type \texttt{Channel};

\item iterated vectorized computation of proper averages of closed vertices;

\item job dequeuing, by recovering finished tasks from a channel and assembling the results into the embedding function $\mathtt{V}: C_0 \to \E^3$ providing an array of type \texttt{Lar.Points} of \texttt{Float64 $\times$ 3}, with vertex coordinates by column.
\end{enumerate}

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{figs/bm_size_tesla.pdf} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\caption{Time requirements of LAR-SURF filter used on artificial with different size of boundary matrix}
\label{fig:bm_size_tesla}
\end{figure}

To compare time requirements of LAR-SURF with Marching Cubes implemented in Python we performed experiment on Ircadb dataset \cite{ircad}. Dataset contain 20 Computed Tomography images (see table \ref{tab:ircad1}) with xy-resolution from 0.56 mm to 0.87 mm and z-resolution from 1.0 mm to 4.0 mm. The number of slices is each series varies from 74 to 260 and size of each slice is $512\times512$. Dataset contain manually segmented liver, portal vein and other structures. We performed surface extraction of liver with Marching Cubes and LAR-SURF. Time required for computation can be seen on figure \ref{fig:ircad_comparison}.

@@

\begin{figure}
\centering
\includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\caption{Time requirements of LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
\label{fig:ircad_comparison}
\end{figure}

>>>>

\section{Julia implementation}\label{sec:julia}

The computer code is implemented in Julia language~\cite{BEKS14} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.

@@

\subsubsection*{Workflow setup}\label{sec:workflow}
The functions in this preliminary step include:
\begin{enumerate}

\item input of 3D medical image $\mathcal{I}$ with \emph{shape} $(\ell_1, \ell_2, \ell_3)$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = [1,2,\ldots,\ell_k]$;

@@

\item depending on the above, best decision for \emph{size} of 3D image block (or brick) $\mathcal{B}$. With defaults $\emph{size}=64$,  the \emph{number of bricks} will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$. 
Hence the default number of bricks is $n=256$, for standard images $512\times 512\times 256$;

\sloppy{
\item computation of the Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the storage of $[\partial_B]$ (for $\emph{size}$ = 64) requires about 45 MB;
}

\item creation of either a local or distributed \texttt{channel} to implement a producer/consumer model of parallel/distributed computation, depending on available resources;

\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (Julia workers) using the Julia macro \texttt{@eveywhere}.

@@

\begin{figure}[tbp]
\includegraphics[width=\textwidth]{figs/schema_horizontal.pdf} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\caption{Workflow of \textsc{lar-surf} algorithm}
\label{fig:schema}
\end{figure}

With $size=64$, the number of non-zeros within the sparse matrix $[\partial(64^3)]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB. The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.

\subsubsection*{Job enqueuing}
% \subsubsection{Job enqueuing}
\label{sec:job-enq}
Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication. Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete. The single work items of this stage follow:
\begin{enumerate}

@@

\subsubsection*{3-Chain encoding}
\label{chain-coding}
The interesting part of the \emph{Image} $\mathcal{I}$ is called \emph{Segment} $\mathcal{S}$. The goal of the whole \emph{workflow} is to extract a \emph{boundary model} of $\mathcal{S}$ from $\mathcal{I}$. The portion of $\mathcal{S}$ inside $\mathcal{B}$, will be denoted as $\mathcal{S}(B)$.
Each block $\mathcal{B}$ of the 3D image must by converted into the \emph{coordinate representation} of a vector $\nu\in C_3$ in the linear space  of 3-chains.

@@

\subsubsection*{SpMM Multiplication}\label{SpMM-multiplication}
According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications. Since we have 256 such jobs (one multiplication per block) to perform in the standard setting of the algorithm\footnote{
Size of block $64^3$; size of image $512^2\times 256$.
}, or more in case of either smaller blocks or image greater than the standard one, this stage must be evidently parallelized and carefully tuned, possibly by using the GPU, if available.
\begin{enumerate}

\item Various multiplication algorithms are being tested, using several packages for sparse linear algebra and/or custom implementations;

@@

\subsubsection*{2-Chain decoding}\label{sec:two-chain-decoding}
Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the \emph{coordinate representation} of the \emph{boundary vector} $\sigma\in C_2$.  The inverse of the coding algorithm is executed in the present stage.  This process can also be partially superimposed in time with the previous ones, depending on the size of the memory buffers used to feed the CPU cores or the GPUs and get their results. Some elementary steps follow:

@@

\subsubsection*{Assembling and artifact filtering}\label{sec:artifact-filtering}
The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the image block $\mathcal{B}$.  In other words, \emph{all quads} of \textbf{each job} are now given in the \textbf{same} \emph{local coordinates}.  Besides to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $\mathcal{B}$ such that $B \cap \mathcal{I} \not= \emptyset$:

@@

\item remove both instances of \emph{double quads} generated by \texttt{Lar} software at the block boundaries 
(see Fig.~\ref{fig:blockboundary}). 
% TODO fix reference probably uncomment next line to link it to Figure 3
% (see Figure~\ref{fig:example}). 
They are artifacts generated by the decomposition of the whole image into a number of blocks of tractable size.

\item 
a smart strategy of removal of such artifacts may be used, which does not require any sorting nor searching on the assembled array of quads. It will consist in arranging each block with all three dimensions decreased-increased by one, so that each 2-adjacent pair of blocks will be covering each other for a full side extent of blocks of depth one. 
% The details of this \emph{artifact filtering} are elucidated in  section~\ref{sec:covering}.
% TODO fix the reference
% TODO is there such a section?

@@

\subsubsection*{Smoothing}\label{sec:smoothing}
The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries. Anyway, the Taubin smoothing~\cite{Taubin1995} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.  Some appropriate set of workers may so be assigned the task of generating iteratively a new position for the vertices they take cure of. In particular, we have:
\begin{enumerate}

\item Job enqueuing, by writing sets of integers (global linear indices of vertices) in array buffers of type \texttt{Channel};

\item vectorized computation of proper averages of near vertices;

\item job dequeuing, by recovering finished tasks from a channel and assembling the results into the embedding function $\mathtt{V}: C_0 \to \E^3$, providing an array of type \texttt{Lar.Points} of \texttt{Float64 $\times$ 3}, with vertex coordinates by column.
\end{enumerate}

@@

\begin{figure}
\centering
\begin{subfigure}{0.495\textwidth}
\includegraphics[width=0.99\textwidth]{figs/bm_size_tesla.pdf} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\subcaption{Time requirements of the \textsc{lar-surf} filter used on artificial volumetric data with various size of boundary matrix}
% \caption{Time requirements of LAR-SURF filter used on artificial with different size of boundary matrix}
\label{fig:bm_size_tesla}
\end{subfigure}
\begin{subfigure}{0.495\textwidth}

\includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\subcaption{Time requirements of the LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
\end{subfigure}
\caption{Performance analysis of the \textsc{lar-surf} filter.}

\end{figure}

To compare time requirements of LAR-SURF with Marching Cubes implemented in Python we performed experiment on Ircadb dataset \cite{ircadb}. Dataset contain 20 Computed Tomography images (see table \ref{tab:ircad2}) with xy-resolution from 0.56 mm to 0.87 mm and z-resolution from 1.0 mm to 4.0 mm. The number of slices is each series varies from 74 to 260 and size of each slice is $512\times512$. Dataset contain manually segmented liver, portal vein and other structures. We performed surface extraction of liver with Marching Cubes and LAR-SURF. Time required for computation can be seen on figure \ref{fig:ircad_comparison}.

@@

% \begin{figure}
% \centering
% \includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
% %\includegraphics[scale=1]{input/ircad_comparison.pdf} 
% \caption{Time requirements of LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
% \label{fig:ircad_comparison}
% \end{figure}
